---
title: "Appendix"
author: "Mukai Wang"
date: "10/28/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(deSolve)
library(dplyr)
rm(list=ls())
```

# Problem 2

For question (a) and (b), I first generate the simulated data using the true $\beta$, $\gamma$ and $\delta$ parameters.


```{r}
initial_state <- c(S=1e6-50, E=0, I=50, R=0)

real_params <- c(beta=5/7, delta=1/3, gamma=1/14, N=1e6)

times <- seq(0, 90, by=1)


SEIR_model <- function(t, state, parameters){
  with(as.list(c(state, parameters)),{
    dS <- -beta * S * I / N
    dE <- beta * S * I / N - delta * E
    dI <- delta * E - gamma * I
    dR <- gamma * I
    list(c(dS, dE, dI, dR))
  })
}

simulated_data <- rk4(y=initial_state,
                      times=times,
                      func = SEIR_model,
                      parms=real_params)

# make sure the counts are integers
simulated_data <- round(simulated_data, digits = 0)
```

The least squares estimation approach has two major components: the search space and the loss function. 

```{r}
beta_space <- seq(0.01, 1, 0.01)

loss.sse <- function(beta){
  params <- c(beta=beta, delta=1/3, gamma=1/14, N=1e6)
  out <- rk4(y=initial_state,
      times=times,
      func = SEIR_model,
      parms=params)
  out <- round(out, digits=0)
  residual <- out[, 'I'] - simulated_data[, 'I']
  sum(residual^2)
}
```


Calculate the loss for all the candidate $\beta$ values and find the one with the smallest SSE.


```{r}
SSE <- sapply(beta_space, loss.sse)
beta_hat <- which.min(SSE)/100
cat(sprintf("The estimated beta is %.2f", beta_hat))
```

For question (c), I add another compartment `OI` to the simulated data to represent the observed infected population size. The observed infected population size is smaller than that in the `I` compartment because of underreporting.

```{r}
initial_state_all <- c(S=1e6-50, E=0, I=50, OI=0,  R=0)
uSEIR_generation_model <- function(t, state, parameters){
  with(as.list(c(state, parameters)),{
    dS <- -beta * S * I / N
    dE <- beta * S * I / N - delta * E
    dI <- delta * E - gamma * I
    dOI <- 0.9 * (delta * E - gamma * I)
    dR <- gamma * I
    list(c(dS, dE, dI, dOI, dR))
  })
}

uSEIR_simulated_data <- rk4(y=initial_state_all,
                      times=times,
                      func = uSEIR_generation_model,
                      parms=real_params)
uSEIR_simulated_data <- round(uSEIR_simulated_data, digits=0)
```

The adjusted SEIR model for estimation only has four compartments. The loss function is the SSE between the `I` compartment in the adjusted SEIR model and the `OI` compartment of the true data.

```{r}
initial_state_observed <- c(S=1e6-50, E=0, I=50,  R=0)

uSEIR_estimation_model <- function(t, state, parameters){
  with(as.list(c(state, parameters)),{
    dS <- - 10 / 9 *beta * S * I / N
    dE <- 10 / 9 * beta * S * I / N - delta * E
    dI <- 9 / 10 * delta * E - gamma * I
    dR <- 1 / 10 * delta * E + gamma * I
    list(c(dS, dE, dI, dR))
  })
}

uSEIR.loss.sse <- function(beta){
  params <- c(beta=beta, delta=1/3, gamma=1/14, N=1e6)
  out <- rk4(y=initial_state_observed,
             times=times,
             func = uSEIR_estimation_model,
             parms=params)
  out <- round(out, digits=0)
  residual <- out[, 'I'] - uSEIR_simulated_data[, 'OI']
  sum(residual^2)
}
```

Calculate the new loss for all the candidate $\beta$ values and report the one with the smallest loss as the estimator.

```{r}
uSEIR_SSE <- sapply(beta_space, uSEIR.loss.sse)
uSEIR_betahat <- which.min(uSEIR_SSE)/100
cat(sprintf("The estimated beta is %.2f", uSEIR_betahat))
```


# Problem 3

I first load in the data and set up the time knots for linear splines of time.

```{r}
folder <- "/home/wangmk/UM/BIOSTAT830/midterm"

michigan_deaths <- readRDS(file.path(folder, 'MI_death_vaccine_combined_data.RDS'))

knot1 <- as.Date("2020-12-14", format="%Y-%m-%d")
knot2 <- as.Date("2021-06-26", format="%Y-%m-%d")
knot3 <- as.Date("2021-11-29", format="%Y-%m-%d")
knot4 <- as.Date("2022-05-22", format="%Y-%m-%d")
vaccine_available <- michigan_deaths$date > knot1
delta <- michigan_deaths$date > knot2
omicron1 <- michigan_deaths$date > knot3
omicron2 <- michigan_deaths$date > knot4

michigan_deaths$vavailable <- as.numeric(vaccine_available)
michigan_deaths$delta <- as.numeric(delta)
michigan_deaths$omicron1 <- as.numeric(omicron1)
michigan_deaths$omicron2 <- as.numeric(omicron2)
```


In question (a), the same translational Poisson log-linear model is fit separately for two age groups
(18-64 and 65+). This is the model 1 in the main text of my submission.

```{r}
adults <- michigan_deaths %>% filter(age == "18-64")
adults$date <- seq(0, 132)
adults_lag1 <- adults[2:133, ]
adults_lag1$deaths <- adults_lag1$deaths + 1
adults_lag1$loglagdeath  <- log(adults$deaths[1:132] + 1)
adults_lag1$totpop <- 6.05e6

model1_adults <- glm(deaths ~ loglagdeath + vaccine_prop_complete + date+
                         I(vavailable * date) +
                         I(delta * date) +
                         I(omicron1 * date) +
                         I(omicron2 * date)  ,
                       family=poisson("log"), data=adults_lag1)
summary(model1_adults)
```



```{r}
senior <- michigan_deaths %>% filter(age == "65+")
senior$date <- seq(0, 132)
senior_lag1 <- senior[2:133, ]
senior_lag1$deaths <- senior_lag1$deaths + 1
senior_lag1$loglagdeath  <- log(senior$deaths[1:132]+1)
senior_lag1$totpop <- 1.81e6

model1_senior <- glm(deaths ~ loglagdeath+
                       vaccine_prop_complete+ date+
                       I(vavailable * date) +
                       I(delta * date) +
                       I(omicron1 * date) +
                       I(omicron2 * date),
                     family=poisson("log"), data=senior_lag1)
summary(model1_senior)
```

For question (b), I combine two age groups and fit model 2 in the main text.

```{r}
combined_data <- rbind(adults_lag1, senior_lag1)
combined_data$agegroup <- as.numeric(combined_data$age == "65+")

model2 <- glm(deaths ~ offset(log(totpop)) + loglagdeath+
                vaccine_prop_complete+ I(agegroup * vaccine_prop_complete)+
                date+
                I(vavailable * date) +
                I(delta * date) +
                I(omicron1 * date) +
                I(omicron2 * date),
              family=poisson("log"), data=combined_data)
summary(model2)
```

For question (c), I prepare two extra columns for the full vaccination rate lags and fit model 3 in the main text.

```{r}
adults_lag2 <- adults_lag1[2:132, ]
adults_lag2$vclag1 <- adults_lag1$vaccine_prop_complete[1:131]
adults_lag2$vclag2 <- c(0, adults_lag1$vaccine_prop_complete[1:130])

senior_lag2 <- senior_lag1[2:132, ]
senior_lag2$vclag1 <- senior_lag1$vaccine_prop_complete[1:131]
senior_lag2$vclag2 <- c(0, senior_lag2$vaccine_prop_complete[1:130])


combined_data2 <- rbind(adults_lag2, senior_lag2)
combined_data2$agegroup <-as.numeric(combined_data2$age == "65+")

model3 <- glm(deaths ~ offset(log(totpop)) + loglagdeath+
                vaccine_prop_complete+ I(agegroup * vaccine_prop_complete)+
                vclag1 + I(agegroup*vclag1)+
                vclag2 + I(agegroup*vclag2)+
                date+
                I(vavailable * date) +
                I(delta * date) +
                I(omicron1 * date) +
                I(omicron2 * date),
              family=poisson("log"), data=combined_data2)
summary(model3)
```
